---
title: "R Notebook"
output: html_notebook
---

# ** NOTE: you should only have to change the file information in lines 41, 43, 46, 48, 51, 53 in the entire script. It is important to follow the formatting guidelines in this chunk, if input data is not correctly formatted this will not work. Everything else should run as is unless you want to change any of the search parameters in the chunk from lines 78-96, some anticipated adjustments are discussed in each line. **

```{r}
## R STUDIO SET UP ##

# install Bioconductor package manager (if not already installed)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.14")
```

```{r}
# define required package list
packages <- c("xlsx", "grDevices", "zoo", "dplyr")
```

```{r}
# install packages if not already installed
BiocManager::install(c(packages))

# respond to "Update all/some/none? [a/s/n]:" popup with "n"
```

```{r}
# load packages every time you relaunch R/R studio
lapply(packages, library, character.only = TRUE)
```

```{r}
## DEFINE FILEPATHS TO THEORETICAL AND EXPERIMENTAL PEAK LISTS AS WELL AS OUTPUT ##
 
# define the filepath to an Excel file containing a fragment list obtained from the UCSC protein prospector MS-Product tool for your protein/PTM combination of interest (https://prospector.ucsf.edu/prospector/cgi-bin/msform.cgi?form=msproduct)
# you want to use the expanded "Theoretical Peak Table" from the bottom of the output screen to generate the Excel file
# make sure that you only have 3 data columns, you will likely need to move things around from how they originally copy over
# additionally, make sure to name your data columns. The exact names aren't important as they will be changed later, but something like m/z, ID, and composition are relevant. The columns must also be in this order, which is how it is originally output from Protein Prospector
theor_FP <- "/Users/smd6459/local_data/MALDI/H3(1-20)_MS2_test.xlsx"
# report the sheet name, even if it is just the default name Sheet1
theor_sheet <- "Theor_K14+K18Ace"

# define the filepath to an Excel file containing the MS2 fragmentation data collected from your protein/PTM sample, this should have 6 columns ordered m/z, S/N, Quality. Fac., Res., Intens., Area. Again the names are not important as they will be changed later
exp_FP <- "/Users/smd6459/local_data/MALDI/H3(1-20)_MS2_test.xlsx"
# report the sheet name, even if it is just the default name Sheet1
exp_sheet <- "Exp_2Ace"

# define the filepath to an Excel file where you want the assigned peak list to go
result_FP <- "/Users/smd6459/local_data/MALDI/H3(1-20)_MS2_test.xlsx"
# define the sheet name where you want your output data stored, this sheet should not already exist or it will throw an error
result_sheet <- "K14+K18Ace_assignment"
```

```{r}
## IMPORT PEAK LISTS FROM EXCEL AND WRANGLE ##

# read in peak lists
theor_fragments <- read.xlsx(theor_FP, sheetName = theor_sheet)
exp_fragments <- read.xlsx(exp_FP, sheetName = exp_sheet)

# rename data columns
colnames(theor_fragments) <- c("m.z", "ID", "Composition")
colnames(exp_fragments) <- c("m.z", "S.N", "Quality", "Res", "Intensity", "Area")
# create ID column in exp_fragments
exp_fragments[ , "ID"] <- NA

# change ID column types from factor to character
theor_fragments$ID <- as.character(theor_fragments$ID)
exp_fragments$ID <- as.character(exp_fragments$ID)

# remove redundant peak(s) if necessary
exp_fragments <- distinct(exp_fragments, m.z, .keep_all = TRUE)
theor_fragments <- distinct(theor_fragments, m.z, .keep_all = TRUE)
```

```{r}
## MATCH EXPERIMENTAL PEAKS TO THEORETICAL PEAKS AND LABEL WITH ION ID ##

# for each theoretical fragment m.z:
for (peak in theor_fragments$m.z) {
# generate a list of experimental m.z that match the theoretical peak within a 2 m.z range and with a quality score greater than 100. This is one of the parameter sets you can change. If you want to change the search range, change -2, 2 to a wider or narrower interval, and if you want to change the quality cutoff change 100 to a more or less discriminating threshold
matches <- which(findInterval(exp_fragments$m.z, c(-2, 2)+peak)==1 & exp_fragments$Quality > 100)
# choose the match with the largest intensity as the best match. If you want to rewrite this to depend on a different variable, change the column number (5) which corresponds to the intensity column as the column number corresponding to your variable or choice
if (length(matches)>0) {
  bestmatch <- max(exp_fragments[matches,5])
# find the best match index in the experimental data. Don't change this
  bestmatchindex <- which(findInterval(exp_fragments$m.z, c(-2, 2)+peak)==1 & exp_fragments$Intensity == bestmatch)
# find the corresponding index in the theoretical data. Don't change this
  peakindex <- which(theor_fragments$m.z == peak)
# replace the empty experimental ID with the ID from the theoretical peak list. Don't change this
  exp_fragments[bestmatchindex,7] <- theor_fragments[peakindex,2]
  }
}
```

```{r}
## OUTPUT RESULTS TO NEW EXCEL LOCATION ##

# write assignments back to an Excel file, when you reopen the file they will be there in a new sheet
write.xlsx(exp_fragments, result_FP, sheetName = result_sheet, 
  col.names = TRUE, row.names = TRUE, append = TRUE)
```
